---
title: "Assignment 2: Coding Basics"
author: "Cynthia Zhou"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. I used "seq" function to create a value named "data", which include numbers from 1 to 30, increasing by threes.
data<-seq(1,30,3)

#2. I used "mean" and "median" function.
mean_d<-mean(data)
median_d<-median(data)

#3. The result is FALSE, indicating the mean is not greater than the median.
mean_d>median_d
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5 create vectors
students <- c("A","B","C","D")
scores <- c(88,92,100,45)
pass <- scores >= 50

#6 I used typeof function to check types of these vectors. students: character, scores: double, pass: logical
typeof(students)
typeof(scores)
typeof(pass)

#7&8
student_grade <- data.frame(student_name = students, test_score = scores,pass_or_not = pass)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: The types of values in data frame can be different while these in a matrix should be same. The columns in data frame have names while there is no clear column name in a matrix.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#option1: `if` and `else` with `for`
check1 <- function(scores) {
  for (x in scores) {
    if (x > 50) {
      print(TRUE)
    } else {
      print(FALSE)
    }
  }
}
check1(scores)

#option2: `ifelse`
check2 <- function(x){
 ifelse(x > 50, TRUE, FALSE)
}
check2(scores)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Both options worked but when using `if` and `else` fuction, `for` should be introduced (credit to ChatGPT). Otherwise, there will be an error: Error in if (x > 50) { : the condition has length > 1. This means "x > 50" directly compares the entire vector with 50, producing a logical vector of the same length as scores.
