---
title: "Assignment 10: Data Scraping"
author: "Cynthia Zhou"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
#Install familiar packages
library(tidyverse);library(lubridate);library(viridis);library(here)

#install.packages("rvest")
library(rvest)

#install.packages("dataRetrieval")
library(dataRetrieval)

#install.packages("tidycensus")
library(tidycensus)

getwd()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
URL <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')
URL
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
system_name <- URL %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
system_name

PWSID <- URL %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
PWSID

Ownership <- URL %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
Ownership

max_dayuse <- URL %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max_dayuse

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data, message=FALSE}
#4 
Month_order <- URL %>% 
  html_nodes(".fancy-table:nth-child(31) tr+ tr th") %>% 
  html_text()
Month_order

df_mdu <- data.frame("Year" = rep(2022,12),
                     "max_daily_withdrawals" = as.numeric(max_dayuse))

df_mdu <- df_mdu %>% 
  mutate(
    Month = !!Month_order,
    System_name = !!system_name,
    PWSID = !!PWSID,
    Ownership = !!Ownership,
    Date = my(paste(Month,"-",Year))) %>% 
  arrange(Date)

glimpse(df_mdu)

#5 
ggplot(df_mdu, aes(x=Date, y=max_daily_withdrawals))+ 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2022 Monthly Water Usage Data for", system_name),
       y="Withdrawal (mgd)",
       x="Date")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
#Create our scraping function
scrape.it <- function(PWSID, the_year){
  
  #Retrieve the website contents 
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
                                  PWSID,'&year=',the_year))
  
  #Set the element address variables (determined in the previous step)
  system_name_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  PWSID_tag <- 'td tr:nth-child(1) td:nth-child(5)'
  Ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  the_data_tag <- 'th~ td+ td'
  
  #Scrape the data items
  system_name <- the_website %>% html_nodes(system_name_tag) %>% html_text()
  PWSID <- the_website %>%   html_nodes(PWSID_tag) %>%  html_text()
  Ownership <- the_website %>% html_nodes(Ownership_tag) %>% html_text()
  max_dayuse <- the_website %>% html_nodes(the_data_tag) %>% html_text()
  
  
  #Convert to a dataframe
  df_withdrawals <- data.frame("Year" = rep(the_year, 12),
                     "max_daily_withdrawals" = as.numeric(max_dayuse))

  df_withdrawals <- df_withdrawals %>% 
    mutate(
      Month = !!Month_order,
      System_name = !!system_name,
      PWSID = !!PWSID,
      Ownership = !!Ownership,
      Date = my(paste(Month,"-",Year))) %>% 
    arrange(Date)
  
  #Pause for a moment - scraping etiquette
  #Sys.sleep(1) #uncomment this if you are doing bulk scraping!
  
  #Return the dataframe
  return(df_withdrawals)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data, message=FALSE}
#7 
the_durham2015 <- scrape.it('03-32-010', 2015)
glimpse(the_durham2015)

ggplot(the_durham2015, aes(x=Date, y=max_daily_withdrawals))+ 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = paste("2015 Monthly Water Usage Data for", system_name),
       y="Withdrawal (mgd)",
       x="Date")
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data, message=FALSE}
#8 
#Asheville data
the_ash2015 <- scrape.it('01-11-010', 2015)

#combine data
df_2015 <- bind_rows(the_durham2015, the_ash2015)
glimpse(df_2015)

#plot
ggplot(df_2015,aes(x=Date,y=max_daily_withdrawals,color=System_name)) + 
  geom_line() + 
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "2015 Water usage data",
       y="Withdrawal (mgd)",
       x="Date")
```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data, message=FALSE}
#9 
the_year_ash <- c(2010:2021)
df_ash2 <- map2('01-11-010', the_year_ash, scrape.it) 
df_ash2 <- bind_rows(df_ash2)

#Plot
ggplot(df_ash2,aes(y = max_daily_withdrawals, x=Date)) +
  geom_line()+
  geom_smooth(method="loess",se=FALSE) +
  labs(title = "Water usage data in Ashveille from 2010 to 2021",
       y="Withdrawal (mgd)",
       x="Date")
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?

> Answer:
> Asheveille had an overall increasing trend in water usage over time among these years. It encountered a slight decrease from 2010 to 2015 but increased after 2015.