---
title: "Assignment 3: Data Exploration"
author: "Cynthia Zhou"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}
getwd()
library(here)
library(readr)
library(tidyverse)
library(lubridate)

Neonics<-read.csv(here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),stringsAsFactors = TRUE) 
Litter<- read.csv(here("Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: 

> Neonicotinoids can affect the central nervous system of insects, leading to their death. While they are widely used in agriculture, some negative effects on food chain or local ecosystem may appear. For example, some beneficial insects or even bigger organisms can be killed that may result in ecological damage. To explore the ecotoxicology of neonicotinoids on insects can help us understand more about this kind of insecticides, and learn how to control and monitor when using them.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: 

> The forest litter and woody debris can be decomposed by some microorganisms, playing an important role in the nutrient cycling, especially carbon cycling. These decomposition can be considered as a crucial carbon sequestration/carbon sink. If they can be buried for a long time, they will benefit in mitigating climate change. Also, they are important in biodiversity, ecosystem services, ect.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 

> 1. Litter and fine woody debris are collected from elevated traps and ground traps, respectively. 

> 2. Sampling is executed in tower plots within the terrestrial NEON sites that contain woody vegetation >2m tall. The number of trap pairs deployed per plot area is determined based on the plot size, with a maximum of 4 trap pairs per plot.

> 3. Ground traps are sampled once per year. Target sampling frequency for elevated traps varies by vegetation present at the site.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r Q6}
summary(Neonics$Effect)
```

> Answer:

> Q5. The dimensions of the dataset is 4623 rows * 30 columns.

> Q6. Population and Mortality are the most common effects. Population and Mortality are the most direct indicators of the effectiveness of neonicotinoids, and the higher the mortality rate, the more effective the insecticide. These two effects show similar results when determining the insects alive or dead.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r Q7}
summary_result<-summary(Neonics$Species.Common.Name)
sorted_summary<-sort(summary_result, decreasing=TRUE)
head(sorted_summary, n=7)
```

> Answer: 

> 1. The six most commonly studied species: Honey Bee(670), Parasitic Wasp Buff(285), Tailed Bumblebee(183), Carniolan Honey Bee(152), Bumble Bee(140), Italian Honeybee(113).

> 2. All these species are golbal distributed, with a large popuation. They are all sensitive to Neonicotinoids and have colonial behavior. They  function as pollinators and contribute to the biodiversity of ecosystems. Because of their wide ditribution, it is easy to conduct and repeat tests all over the world. Their colonial behavior helps to study the effects of insecticides at both individual and colony levels. And being as pollinators in the ecosystem makes it important to learn the impacts of Neonicotinoids on them.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r Q8}
class(Neonics$Conc.1..Author.)
```

> Answer: It is factor. Because it includes symbol "/" rather than only numbers.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Q9}
ggplot(Neonics)+
  geom_freqpoly(aes(x=Publication.Year))+
  labs(title="Number of Studies Conducted by Publication Year", x="Publication Year", y="Count")
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Q10}
ggplot(Neonics)+
  geom_freqpoly(aes(x=Publication.Year,color=Test.Location))+  
  labs(title="Number of Studies Conducted by Publication Year", x="Publication Year", y="Count")
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: 

> Lab and Natural Field are the most common locations. Both of them have increased from a very low frequency since 1990, showing a fluctuating trend. Between 2005 and 2010, Natural Field are more common than Lab, while Lab laid on the dominant position during other periods.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}
ggplot(Neonics)+
  geom_bar(aes(x=Endpoint))+
  labs(title="Number of End Points", x="End Points", y="Count")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

> Answer: 

> NOEL and LOEL.

> NOEL(No-observable-effect-level): highest dose (concentration) producing effects not significantly different from responses of controls according to author's reported statistical test.

> LOEL(Lowest-observable-effect-level): lowest dose (concentration) producing effects that were significantly different (as reported by authors) from responses of controls.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}
class(Litter$collectDate) # It's a factor not a date
Litter$collectDate<-ymd(Litter$collectDate)
class(Litter$collectDate) # Now, it's a date

unique(Litter$collectDate[format(Litter$collectDate,"%Y-%m")=="2018-08"])
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}
unique(Litter$namedLocation)
summary(Litter$namedLocation)
```

> Answer:

> Q12. "2018-08-02" and "2018-08-30"

> Q13. 12 plots were sampled. `unique` function outputs a list of unique values. `Summary` function counts each values respectively.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}
ggplot(Litter)+
  geom_bar(aes(x=functionalGroup))+
  labs(title="Number of Functional Group", x="Functional Group", y="Count")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Q15}
ggplot(Litter)+
  geom_boxplot(aes(x=functionalGroup,y=dryMass))+
  labs(title="Dry Biomass in Different Types of Litter ", x="Functional Group", y="Bry Biomass")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(Litter)+
  geom_violin(aes(x=functionalGroup,y=dryMass))+
  labs(title="Dry Biomass in Different Types of Litter ", x="Functional Group", y="Bry Biomass")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: 

> The boxplot provides more inforamtion here, clearly showing the median, quartiles, and potential outliers in the data. The violin plot provides only the values of biomass.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: 

> Needles have the highest biomass, followed by Twigs/branches and Mixed.